maze-dataset

### 2025-03-30 docs/paper/paper.tex
paper wip

\begin{figure} 
  \begin{minipage}{5in}
    \input{diagram/diagram.tikz} 
  \end{minipage}
  \caption{Usage of maze-dataset. We create a `MazeDataset` from a `MazeDatasetConfig`. This contains `SolvedMaze` objects which can be converted to and from a variety of formats. Code in the image contains clickable links to \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html}{documentation}. A variety of generated examples can be viewed \href{https://understanding-search.github.io/maze-dataset/examples/maze_examples.html}{here}.}
  \label{fig:diagram}
\end{figure}

@@

When initializing mazes, further configuration options can be specified
through the \texttt{from\_config()} factory method as necessary. Options
allow for saving/loading existing datasets instead of regenerating, and
parallelization options for generation. Available maze generation
algorithms are static methods of the \texttt{LatticeMazeGenerators}
class and include the following:

\begin{itemize}
\tightlist
\item
  \texttt{gen\_dfs} \textbf{(randomized depth-first search)}: Parameters
  can be passed to constrain the number of accessible cells, the number
  of forks in the maze, and the maximum tree depth. Creates a spanning
  tree by default or a partially spanning tree if constrained.
\item
  \texttt{gen\_wilson} \textbf{(Wilson's algorithm)}: Generates a random
  spanning tree via loop-erased random walk
  (\protect\hyperlink{ref-wilson}{Wilson, 1996}).
\item
  \texttt{gen\_percolation} \textbf{(percolation)}: Starting with no
  connections, every possible lattice connection is set to either true
  or false with some probability \(p\), independently of all other
  connections. For the kinds of graphs that this process generates, we
  refer to existing work
  (\protect\hyperlink{ref-percolation}{Duminil-Copin, 2017};
  \protect\hyperlink{ref-percolation-clustersize}{Fisher \& Essam,
  2004}).
\item
  \texttt{gen\_dfs\_percolation} \textbf{(randomized depth-first search
  with percolation)}: A connection exists if it exists in a maze
  generated via \texttt{gen\_dfs\ OR\ gen\_percolation}. Useful for
  generating mazes that are not acyclic graphs.
\end{itemize}

Furthermore, a dataset of mazes can be filtered to satisfy certain
properties:

@@

Custom filters can be specified, and several filters are included:

\begin{itemize}
\tightlist
\item
  \texttt{path\_length(min\_length:\ int)}: shortest length from the
  origin to target should be at least \texttt{min\_length}.
\item
  \texttt{start\_end\_distance(min\_distance:\ int)}: Manhattan distance
  between start and end should be at least \texttt{min\_distance},
  ignoring walls.
\item
  \texttt{remove\_duplicates(...)}: remove mazes which are similar to
  others in the dataset, measured via Hamming distance.
\item
  \texttt{remove\_duplicates\_fast()}: remove mazes which are exactly
  identical to others in the dataset.
\end{itemize}

@@

Internally, mazes are \texttt{SolvedMaze} objects, which have path
information, and a connection list optimized for storing sub-graphs of a
lattice. These objects can be converted to and from several formats.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3250}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\texttt{as\_ascii()}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\texttt{as\_pixels()}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\texttt{MazePlot()}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Simple text format for displaying mazes, useful for debugging in a
terminal environment. & \texttt{numpy} array of \texttt{dtype=uint8} and
shape \texttt{(height,\ width,\ 3)}. The last dimension is RGB color. &
feature-rich plotting utility with support for multiple paths, heatmaps
over positions, and more. \\
\end{longtable}

\begin{figure}
\hypertarget{fig:visualoutputs}{%
\centering
\includegraphics{figures/output-fmts.pdf}
\caption{Various output formats. Top row (left to right): ASCII diagram,
rasterized pixel grid, and advanced display.}\label{fig:visualoutputs}
}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{docs/benchmarks/figures/n_mazes-vs-gentime.pdf}
\caption{Maze size vs generation time}
\end{figure}

@@

Parallelization is implemented via the \texttt{multiprocessing} module
in the Python standard library, and parallel generation can be
controlled via keyword arguments to the
\texttt{MazeDataset.from\_config()} function.

@@

The
\href{https://github.com/understanding-search/maze-dataset}{\texttt{maze-dataset}}
library (\protect\hyperlink{ref-maze-dataset-github}{Michael I.
Ivanitskiy et al., 2023a}) introduced in this paper provides a flexible
and extensible toolkit for generating, processing, and analyzing maze
datasets. By supporting various procedural generation algorithms and
conversion utilities, it enables the creation of mazes with customizable
properties to suit diverse research needs. Planned improvements to the
\texttt{maze-dataset} include adding more generation algorithms (such as
Prim's algorithm (\protect\hyperlink{ref-dijkstra-prim}{Dijkstra, 1959};
\protect\hyperlink{ref-jarnik-prim}{Jarnık, 1930};
\protect\hyperlink{ref-prim}{Prim, 1957}) and Kruskal's algorithm
(\protect\hyperlink{ref-kruskal}{Kruskal, 1956}), among others
(\protect\hyperlink{ref-mazegen_analysis}{Gabrovšek, 2019})), adding the
ability to augment a maze with an adjacency list to add ``shortcuts'' to
the maze, and resolving certain limitations detailed in Section
\protect\hyperlink{limitations}{Limitations}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kruskal}{}}%
Kruskal, J. B. (1956). On the shortest spanning subtree of a graph and
the traveling salesman problem. \emph{Proceedings of the American
Mathematical Society}, \emph{7}(1), 48--50.
\url{https://doi.org/10.1090/S0002-9939-1956-0078686-7}

>>>>

\begin{figure} 
  \begin{minipage}{5in}
    \input{diagram/diagram.tikz} 
  \end{minipage}
  \caption{
    Usage of maze-dataset. We create a \texttt{MazeDataset} from a \texttt{MazeDatasetConfig}. This contains \texttt{SolvedMaze} objects which can be converted to and from a variety of formats. Code in the image contains clickable links to \href{https://understanding-search.github.io/maze-dataset/maze_dataset.html}{documentation}. A variety of generated examples can be viewed \href{https://understanding-search.github.io/maze-dataset/examples/maze_examples.html}{here}.
  }
  \label{fig:diagram}
\end{figure}

@@

When initializing mazes, further configuration options can be specified
through the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.from_config}{\texttt{from\_config()}}
factory method as necessary. Options allow for saving/loading existing
datasets instead of regenerating, and parallelization options for
generation. Available maze generation algorithms are static methods of
the \texttt{LatticeMazeGenerators} class and include generation
algorithms based on randomized depth-first search, Wilson's algorithm
(\protect\hyperlink{ref-wilson}{Wilson, 1996}), percolation
(\protect\hyperlink{ref-percolation}{Duminil-Copin, 2017};
\protect\hyperlink{ref-percolation-clustersize}{Fisher \& Essam, 2004}),
Kruskal's algorithm
(\protect\hyperlink{ref-kruskal1956shortest}{Kruskal, 1956a}), and
others.

Furthermore, a dataset of mazes can be filtered to satisfy certain
properties. Custom filters can be specified, and some filters are
included in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/filters.html\#MazeDatasetFilters}{\texttt{MazeDatasetFilters}}.

@@

@@

Internally, mazes are
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#SolvedMaze}{\texttt{SolvedMaze}}
objects, which have path information, and a connection list optimized
for storing sub-graphs of a lattice. These objects can be converted to
and from several formats.

\begin{figure}[H]
  \centering
  \begin{tabular}{p{2in} p{2in} p{2in}} 
    \hline \\[.5em]
    % algorithms
    \texttt{as\_ascii()} & \texttt{as\_pixels()} & \texttt{MazePlot()} \\[.5em]
    % descriptions
      Simple text format for displaying mazes, useful for debugging in a terminal environment.
      & \texttt{numpy} array of \texttt{dtype=uint8} and shape \texttt{(height, width, 3)}. The last dimension is RGB color.
      & feature-rich plotting utility with support for multiple paths, heatmaps over positions, and more. \\[1em]
    \hline \\
    % examples
      \multicolumn{1}{c}{\begin{minipage}[b]{1.6in}
        \input{figures/outputs-ascii-colored.tex} 
      \end{minipage}}
      & \multicolumn{1}{c}{
        \includegraphics[width=0.25\textwidth]{figures/outputs-pixels.pdf}
      }
      & \multicolumn{1}{c}{
        \includegraphics[width=0.25\textwidth, trim={0 0 -.3cm, -.5cm}, clip]{figures/outputs-mazeplot.pdf}
      } \\[1em]

    \hline \\
  \end{tabular}
  \caption{Various output formats. Top row (left to right): ASCII diagram, rasterized pixel grid, and advanced display.}
  \label{fig:output-fmts}
\end{figure}

@@

@@

Parallelization is implemented via the \texttt{multiprocessing} module
in the Python standard library, and parallel generation can be
controlled via keyword arguments to the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.from_config}{\texttt{MazeDataset.from\_config()}}
function.

@@

The
\href{https://github.com/understanding-search/maze-dataset}{\texttt{maze-dataset}}
library (\protect\hyperlink{ref-maze-dataset-github}{Michael I.
Ivanitskiy et al., 2023a}) introduced in this paper provides a flexible
and extensible toolkit for generating, processing, and analyzing maze
datasets. By supporting various procedural generation algorithms and
conversion utilities, it enables the creation of mazes with customizable
properties to suit diverse research needs. Planned improvements to the
\texttt{maze-dataset} include adding more generation algorithms (such as
Prim's algorithm (\protect\hyperlink{ref-dijkstra-prim}{Dijkstra, 1959};
\protect\hyperlink{ref-jarnik-prim}{Jarnık, 1930};
\protect\hyperlink{ref-prim}{Prim, 1957}) and Kruskal's algorithm
(\protect\hyperlink{ref-kruskal}{Kruskal, 1956b}), among others
(\protect\hyperlink{ref-mazegen_analysis}{Gabrovšek, 2019})), adding the
ability to augment a maze with an adjacency list to add ``shortcuts'' to
the maze, and resolving certain limitations detailed in Section
\protect\hyperlink{limitations}{Limitations}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kruskal1956shortest}{}}%
Kruskal, J. B. (1956a). On the shortest spanning subtree of a graph and
the traveling salesman problem. \emph{Proceedings of the American
Mathematical Society}, \emph{7}(1), 48--50.

\leavevmode\vadjust pre{\hypertarget{ref-kruskal}{}}%
Kruskal, J. B. (1956b). On the shortest spanning subtree of a graph and
the traveling salesman problem. \emph{Proceedings of the American
Mathematical Society}, \emph{7}(1), 48--50.
\url{https://doi.org/10.1090/S0002-9939-1956-0078686-7}
